The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes

### 2024-12-15 draft_finalreport.qmd
updating final report

@@

```{r, echo=FALSE, results=FALSE}
# Data Cleaning and Recoding for dataset_new
dataset_clean <- dataset_new |> 
  mutate(
    # Education recoding
    educ = case_when(
      S13 == "(1) Grades 1-8" ~ "No college experience",
      S13 == "(2) Some High School, but did not graduate" ~ "No college experience",
      S13 == "(3) High School graduate or GED" ~ "No college experience",
      S13 == "(4) Some college" ~ "Some college experience",
      S13 == "(5) Associates, 2-year degree" ~ "College graduate",
      S13 == "(6) Bachelors, 4-year degree" ~ "College graduate",
      S13 == "(7) Post-graduate degree" ~ "College graduate"
    ),
    # Sexual orientation recoding
    sex_orientation = case_when(
      S3 == "(01) Straight, that is, not gay or lesbian" ~ "Straight",
      S3 == "(02) Gay or lesbian" ~ "Queer",
      S3 == "(03) Bisexual" ~ "Queer",
      S3 == "(04) Something else (Specify)" ~ "Queer",
      S3 == "(88) Refused" ~ "Refused/Don't Know",
      S3 == "(99) I don" ~ "Refused/Don't Know"
    ),
    # Urban vs rural recoding
    rural = case_when(
      S14 == "(1) Large urban area" ~ "Urban/suburban",
      S14 == "(2) Large suburb near large city" ~ "Urban/suburban",
      S14 == "(3) Small suburb near small town or city" ~ "Urban/suburban",
      S14 == "(4) Small town or small city" ~ "Urban/suburban",
      S14 == "(5) Rural area" ~ "Rural"
    ),
    # Race recoding
    race_recoded = case_when(
      S2_RACE_PRIME == "(1) White" ~ "White",
      S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Latino",
      S2_RACE_PRIME == "(3) Black or African American" ~ "Black",
      S2_RACE_PRIME == "(4) Asian American" ~ "Asian American",
      S2_RACE_PRIME == "(5) American Indian/Native American" ~ "AIAN",
      S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Arab/MENA",
      S2_RACE_PRIME == "(7) Native Hawaiian" ~ "NHPI",
      S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "NHPI"
    ),
    # Age recoding
    age_recoded = case_when(
      S5_AGE == "(1) 16-17" ~ "16-17",
      S5_AGE == "(2) 18-29" ~ "18-29",
      S5_AGE == "(3) 30-39" ~ "30-39",
      S5_AGE == "(4) 40-49" ~ "40-49",
      S5_AGE == "(5) 50-59" ~ "50-59",
      S5_AGE == "(6) 60-69" ~ "60-69",
      S5_AGE == "(7) 70 +" ~ "70+"
    )
  ) |>
  mutate(
    # Converting variables to factors
    race_recoded = factor(race_recoded, levels = c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA")),
    age_recoded = factor(age_recoded, levels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")),
    educ = factor(educ, levels = c("No college experience", "Some college experience", "College graduate")),
    rural = factor(rural, levels = c("Rural", "Urban/suburban")),
    sex_orientation = factor(sex_orientation, levels = c("Straight", "Queer", "Refused/Don't Know"))
  )

# Take a look at the updated dataset
head(dataset_clean)

@@

dataset_clean <- dataset_clean[, !(names(dataset_clean) %in% redundant_columns)]
```

@@

## VIF Table - Research Question 2

@@

@@

## Table MODEL: Base Model Statistics
|                           | Metric  |
|---------------------------|---------|
| Sensitivity (Recall)      | 0.1563  |
| Specificity               | 0.9489  |
| Precision                 | 0.5146  |
| F1                        | 0.2397  |
| Balanced Accuracy         | 0.5526  |
| No Information Rate (NIR) | 0.7425  |
| Kappa                     | 0.1361  |
| Mcnemar's Test P-Value    | \<2e-16 |

>>>>

  # Outcome Variable: Level of Political Interest
  "Q29",

  # Predictor: Importance of Racial Belonging
  "Q271",

  # Predictor: Party ID
  "Q21",

@@

```{r, echo=FALSE, results=FALSE}
# Data Cleaning and Recoding for dataset_new
dataset_clean_allvars <- dataset_new |> 
  mutate(
    # Education recoding
    educ = case_when(
      S13 == "(1) Grades 1-8" ~ "No college experience",
      S13 == "(2) Some High School, but did not graduate" ~ "No college experience",
      S13 == "(3) High School graduate or GED" ~ "No college experience",
      S13 == "(4) Some college" ~ "Some college experience",
      S13 == "(5) Associates, 2-year degree" ~ "College graduate",
      S13 == "(6) Bachelors, 4-year degree" ~ "College graduate",
      S13 == "(7) Post-graduate degree" ~ "College graduate"
    ),
    # Sexual orientation recoding
    sex_orientation = case_when(
      S3 == "(01) Straight, that is, not gay or lesbian" ~ "Straight",
      S3 == "(02) Gay or lesbian" ~ "Queer",
      S3 == "(03) Bisexual" ~ "Queer",
      S3 == "(04) Something else (Specify)" ~ "Queer",
      S3 == "(88) Refused" ~ "Refused/Don't Know",
      S3 == "(99) I don" ~ "Refused/Don't Know"
    ),
    # Urban vs rural recoding
    rural = case_when(
      S14 == "(1) Large urban area" ~ "Urban/suburban",
      S14 == "(2) Large suburb near large city" ~ "Urban/suburban",
      S14 == "(3) Small suburb near small town or city" ~ "Urban/suburban",
      S14 == "(4) Small town or small city" ~ "Urban/suburban",
      S14 == "(5) Rural area" ~ "Rural"
    ),
    # Race recoding
    race_recoded = case_when(
      S2_RACE_PRIME == "(1) White" ~ "White",
      S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Latino",
      S2_RACE_PRIME == "(3) Black or African American" ~ "Black",
      S2_RACE_PRIME == "(4) Asian American" ~ "Asian American",
      S2_RACE_PRIME == "(5) American Indian/Native American" ~ "AIAN",
      S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Arab/MENA",
      S2_RACE_PRIME == "(7) Native Hawaiian" ~ "NHPI",
      S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "NHPI"
    ),
    # Age recoding
    age_recoded = case_when(
      S5_AGE == "(1) 16-17" ~ "16-17",
      S5_AGE == "(2) 18-29" ~ "18-29",
      S5_AGE == "(3) 30-39" ~ "30-39",
      S5_AGE == "(4) 40-49" ~ "40-49",
      S5_AGE == "(5) 50-59" ~ "50-59",
      S5_AGE == "(6) 60-69" ~ "60-69",
      S5_AGE == "(7) 70 +" ~ "70+"
    )
  ) |>
  mutate(
    # Converting variables to factors
    race_recoded = factor(race_recoded, levels = c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA")),
    age_recoded = factor(age_recoded, levels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")),
    educ = factor(educ, levels = c("No college experience", "Some college experience", "College graduate")),
    rural = factor(rural, levels = c("Rural", "Urban/suburban")),
    sex_orientation = factor(sex_orientation, levels = c("Straight", "Queer", "Refused/Don't Know"))
  )

@@

dataset_clean <- dataset_clean_allvars[, !(names(dataset_clean_allvars) %in% redundant_columns)]
```

```{r echo=FALSE, results=FALSE}
# Rename remaining columns in dataset_clean
colnames(dataset_clean) <- c(
  "ImportanceOfStoppingDiscriminationAgainstMinorities",  # Q1R5
  "Gender",                                               # S3B
  "TreatedUnfairlyOrDiscriminatedAgainst",                # Q627
  "DiscriminationImpactOnLifeSatisfaction",              # Q633
  "PerceptionsOfDiscriminationAgainstWhites",            # Q619_Q626R1
  "PerceptionsOfDiscriminationAgainstBlacks",            # Q619_Q626R2
  "PerceptionsOfDiscriminationAgainstAsianAmericans",    # Q619_Q626R3
  "PerceptionsOfDiscriminationAgainstNativeAmericans",   # Q619_Q626R4
  "PerceptionsOfDiscriminationAgainstImmigrants",        # Q619_Q626R5
  "PerceptionsOfDiscriminationAgainstLatinos",           # Q619_Q626R6
  "PerceptionsOfDiscriminationAgainstGaysAndLesbians",   # Q619_Q626R7
  "PerceptionsOfDiscriminationAgainstMuslims",           # Q619_Q626R8
  "Education",                                           # educ
  "SexOrientation",                                      # sex_orientation
  "Rural",                                               # rural
  "Race",                                                # race_recoded
  "Age"                                                  # age_recoded
)

```{r echo=FALSE, results=FALSE}

## Calculating Odds Ratio, Confidence Interval, and p-values (base model)

# Extract odds ratios
odds_ratios <- exp(coef(base_model))

# Get confidence intervals for coefficients
conf_intervals <- exp(confint(base_model))

# Extract p-values
p_values <- summary(base_model)$coefficients[, 4]

# Combine everything into a single data frame
results_base_model <- data.frame(
  OddsRatio = odds_ratios,
  LowerCI = conf_intervals[, 1],
  UpperCI = conf_intervals[, 2],
  PValue = p_values
)

# Display results
results_base_model
```

@@

## Research Question 1

### Model Summary - OLR Model with Interactions
```{r}
dataset_clean_allvars$Q271_reordered <- factor(dataset_clean_allvars$Q271,
                            levels = c("(5) Not at all important",
                                     "(4) Slightly important",
                                     "(3) Moderately important",
                                     "(2) Very important",
                                     "(1) Extremely important"))
dataset_clean_allvars$Q29_reordered <- factor(dataset_clean_allvars$Q29,
                            levels = c("(4) Not at all interested in politics",
                                     "(3) Not that interested in politics",
                                     "(2) Somewhat interested",
                                     "(1) Very interested in politics"), ordered=TRUE)

ord_mod_interactions <- polr(Q29_reordered ~ Q271_reordered * S3B +
                               race_recoded +
                  age_recoded +
            educ +  # Educational Attainment
            Q21, # Party affiliation
          data = dataset_clean_allvars, Hess=TRUE)

summary(ord_mod_interactions)

pvals_interact <- pnorm(-abs(summary(ord_mod_interactions)$coef[,"t value"]))*2
ctable_interact <- cbind(summary(ord_mod_interactions)$coef,pvals_interact)

ctable_interact

(exp_coefs_interact <- exp(cbind(OR=coef(ord_mod_interactions),confint(ord_mod_interactions))))

head(ord_mod_interactions$fitted.values)
head(predict(ord_mod_interactions))

confusionMatrix(predict(ord_mod_interactions), dataset_clean$Q29, mode="everything")

### Odds Ratio Table 
| Variable | Odds Ratio | 2.5% (Confidence Interval) | 97.5% (Confidence Interval) |
|------------------|------------------|------------------|------------------|
| Q271 (4) Slightly important | 0.9672589 | 0.811845068 | 1.1525850 |
| Q271 (3) Moderately important | 0.9776608 | 0.845556718 | 1.1303745 |
| Q271 (2) Very important | 1.1952004 | 1.033019394 | 1.3828451 |
| Q271 (1) Extremely important | 1.6474969 | 1.423670208 | 1.9066091 |
| S3B(2) Woman | 0.6294256 | 0.540015144 | 0.7336072 |
| S3B(3) Non-binary | 1.8576658 | 1.002509318 | 3.4617409 |
| S3B(4) Something else (Specify) | 2.1359493 | 0.377740111 | 13.5124525 |
| race Black | 0.8676620 | 0.790376010 | 0.9524806 |
| race Asian American | 0.5945609 | 0.541964870 | 0.6522143 |
| race Latino | 0.7946622 | 0.575235415 | 0.7848896 |
| race AIAN | 0.6719204 | 1.27404185 | 1.7384045 |
| race NHPI | 0.5606006 | 0.458205702 | 0.6860792 |
| race Arab/MENA | 1.1589043 | 0.899447122 | 1.4956228 |
| age 30-39 | 1.0310801 | 0.946696738 | 1.0563119 |
| age 40-49 | 0.9247623 | 0.98549367 | 1.1229868 |
| age 50-59 | 1.0979487 | 1.000027913 | 1.2054892 |
| age 60-69 | 1.3238874 | 1.205240800 | 1.4542874 |
| age 70+ | 1.6896464 | 1.510226543 | 1.8907050 |
| educSome college experience | 1.5264694 | 1.402573135 | 1.6614197 |
| educCollege graduate | 2.2337286 | 2.072162224 | 2.4081069 |
| Q21(2) Democrat | 1.1333513 | 1.045765624 | 1.2282275 |
| Q21(3) Independent | 0.5806653 | 0.533801211 | 0.6315870 |
| Q21(4) Other party | 0.2127592 | 0.182654104 | 0.2477198 |
| Q271 4) Slightly important:S3B(2) Woman | 0.9955364 | 0.782564309 | 1.2663323 |
| Q271 (3) Moderately important:S3B(2) Woman | 0.9461281 | 0.780040309 | 1.1475378 |
| Q271 (2) Very important:S3B(2) Woman | 0.9791675 | 0.810674379 | 1.1826356 |
| Q271 (1) Extremely important:S3B(2) Woman | 0.8114102 | 0.674275440 | 0.9763646 |
| Q271 (4) Slightly important:S3B(3) Non-binary | 0.6505015 | 0.198565161 | 2.1505278 |
| Q271 (3) Moderately important:S3B(3) Non-binary | 0.6563254 | 0.249594245 | 1.7277048 |
| Q271 (2) Very important:S3B(3) Non-binary | 0.5140828 | 0.205132143 | 1.2905898 |
| Q271 (1) Extremely important:S3B(3) Non-binary | 0.8770903 | 0.354410521 | 2.1835545 |
| Q271 (4) Slightly important:S3B(4) Something else (Specify) | 0.2279084 | 0.023857021 | 2.0164000 |
| Q271 (3) Moderately important:S3B(4) Something else (Specify) | 1.7158304 | 0.146865519 | 19.7516328 |
| Q271 2) Very important:S3B(4) Something else (Specify) | 0.1049752 | 0.006213617 | 1.4210951 |
| Q271 1) Extremely important:S3B(4) Something else (Specify) | 0.6584933 | 0.079780177 | 4.9850479 |

## Research Question 2

### VIF Table

@@

### Odds Ratio Table

@@

### Table MODEL: Base Model Statistics
|                           | Metric  |
|---------------------------|---------|
| Sensitivity (Recall)      | 0.1563  |
| Specificity               | 0.9489  |
| Precision                 | 0.5146  |
| F1                        | 0.2397  |
| Balanced Accuracy         | 0.5526  |
| No Information Rate (NIR) | 0.7425  |
| Kappa                     | 0.1361  |
| Mcnemar's Test P-Value    | \<2e-16 |
